NumPy---Array Programming in Python

### 2020-02-20 body.tex
Remove old text

% KJM: I moved this here to get the placement right
% FIXME SJW: let's review caption (d); I don't think it's crystal clear what
% we're trying to show there; double check caption (g)
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth]{static/sketches/concepts}
  \caption{\textbf{The NumPy array incorporates several fundamental array concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with slices and steps. These
  operations return a \emph{view} of the original data.
    \textbf{c,} Indexing an array with masks, scalar coordinates, or
  other arrays, so that it returns a copy of the original data. In the
  bottom example, an array is indexed with other arrays; this broadcasts the indexing arguments before performing the lookup.
    \textbf{d,} Vectorization efficiently applies operations to groups
  of elements.
    \textbf{e,} Broadcasting in the multiplication of two-dimensional arrays.
    \textbf{f,} Reduction operations act along one or more axes. In this
    example, an array is be summed along various axes to produce a vector, or along two axes consecutively to
    produce a scalar.
    \textbf{g,} Example NumPy code, illustrating some of these concepts.
   }
  \label{fig:array-concepts}
\end{figure*}

@@

% this is the software engineering side of things
% as opposed to the interactive, exploration described above

@@

%CuPy, PyData/Sparse, and JAX the most prominent examples---while others have
%NumPy-inspired semantics but, often due to their development history, deviate in places.
%Wherever there are such deviations, we often observe users gently requesting,
%and developers evolving the software step by step toward the NumPy API.

% when they deviate, they sometimes fix it, e.g.,
% https://github.com/pytorch/pytorch/pull/1983
% https://github.com/pytorch/pytorch/issues/2228
% many have for numpy users
% https://github.com/torch/torch7/wiki/Torch-for-Numpy-users

% computation graphs---which postpone and combine calculations
% before executing them efficiently,

@@

% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
%Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:

% should we hint at some of the community interaction part
% they have helped us develop the neps and, e.g.
% https://github.com/pytorch/pytorch/issues/22402

% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
%However, general adoption has not yet been achieved mainly for compatibility
%concerns.
% explain a bit what "compatibility concerns" means ...
%Thus, we are actively addressing these concerns by revising and extending
%these protocols with the aim of wide interoperability that spans the whole
%scientific Python ecosystem up to and including domain libraries.

% https://numpy.org/neps/nep-0016-abstract-array.html
% https://numpy.org/neps/nep-0018-array-function-protocol.html
% https://numpy.org/neps/nep-0022-ndarray-duck-typing-overview.html
% https://numpy.org/neps/nep-0030-duck-array-protocol.html
% https://github.com/numpy/numpy/blob/a111b551ae940d7d5f8523fef1cf3589c6ba00a0/doc/neps/nep-0033-extensible-dtypes.rst
% https://numpy.org/neps/nep-0037-array-module.html

@@

%NumPy has subsequently seen prolific adoption, with millions of users,
%first from research and later also industry.
%It spurred the development of the broader scientific Python ecosystem and
%heralded the current era of widespread use of Python for scientific
%computing. % This and the "What began as" sentence below are more or less interchangeable.

@@

Over the next decade, we will face several new challenges.
%Artificial
%Intelligence and Data Science will mature ... [what do we say about
%  this?].
New devices will be developed, and existing specialized
hardware will evolve, to meet diminishing returns on Moore's law.
There will be more, and a wider variety of, data science
practitioners, a significant proportion of whom will be using NumPy.
The scale of scientific data gathering will continue to expand, with
the adoption of devices and instruments such as light sheet
microscopes and the Large Synoptic Survey Telescope (LSST) \cite{jenness2018lsst}.  New generation languages, interpreters, and compilers,
such as Rust \cite{10.1145/2692956.2663188}, Julia \cite{Julia-2017}, and LLVM \cite{LLVM:CGO04}, will invent
and determine the viability of new concepts and data structures.

>>>>

% KJM: I moved this here to get the placement right
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth]{static/sketches/concepts}
  \caption{\textbf{The NumPy array incorporates several fundamental array concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with slices and steps. These
  operations return a \emph{view} of the original data.
    \textbf{c,} Indexing an array with masks, scalar coordinates, or
  other arrays, so that it returns a copy of the original data. In the
  bottom example, an array is indexed with other arrays; this broadcasts the indexing arguments before performing the lookup.
    \textbf{d,} Vectorization efficiently applies operations to groups
  of elements.
    \textbf{e,} Broadcasting in the multiplication of two-dimensional arrays.
    \textbf{f,} Reduction operations act along one or more axes. In this
    example, an array is be summed along various axes to produce a vector, or along two axes consecutively to
    produce a scalar.
    \textbf{g,} Example NumPy code, illustrating some of these concepts.
   }
  \label{fig:array-concepts}
\end{figure*}

@@

@@

@@

@@

@@

Over the next decade, we will face several new challenges.  New devices will be
developed, and existing specialized hardware will evolve, to meet diminishing
returns on Moore's law.  There will be more, and a wider variety of, data
science practitioners, a significant proportion of whom will be using NumPy.
The scale of scientific data gathering will continue to expand, with the
adoption of devices and instruments such as light sheet microscopes and the
Large Synoptic Survey Telescope (LSST) \cite{jenness2018lsst}.  New generation
languages, interpreters, and compilers, such as Rust
\cite{10.1145/2692956.2663188}, Julia \cite{Julia-2017}, and LLVM
\cite{LLVM:CGO04}, will invent and determine the viability of new concepts and
data structures.

### 2020-02-20 summary.tex
Remove old text

% - 2-3 sentences of basic-level introduction to the field;
Array programming provides a powerful, compact, expressive syntax for accessing,
manipulating, and operating on data in vectors, matrices, and
higher-dimensional arrays \cite{iverson1980notation}.
% - a brief account of the background and rationale of the work;
NumPy is the fundamental array programming library for the Python language
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Developed for over two decades by more than 850 contributors, 
it has nearly 300,000 dependent repositories on GitHub and millions of
downloads per year.
%It is one of the most commonly used machine learning tools among
%enterprises \cite{451report2018}.
It plays an essential role in research analysis pipelines in fields as
diverse as physics, chemistry, astronomy, geoscience, biology, psychology,
material science, engineering, finance, and economics.
For example, in astronomy, NumPy was an important part of the software stack used
in the discovery of gravitational waves \cite{abbott2016observation}
and the first imaging of a black hole \cite{eht-imaging}.
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
Here we show how a few fundamental concepts lead to a simple and
powerful programming paradigm for organizing, exploring, and analyzing
scientific data.
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.
NumPy is the foundation upon which the entire scientific Python
universe is constructed. It is so pervasive that several projects,
targeting audiences with specialized needs, have developed their own
NumPy-like interfaces and array objects.  Because of its central position in the
ecosystem, NumPy increasingly plays the role of an interoperability layer
between these new array computation libraries.

>>>>

% - 2-3 sentences of basic-level introduction to the field;
Array programming provides a powerful, compact, expressive syntax for accessing,
manipulating, and operating on data in vectors, matrices, and
higher-dimensional arrays \cite{iverson1980notation}.
% - a brief account of the background and rationale of the work;
NumPy is the primary array programming library for the Python language
\cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It plays an essential role in research analysis pipelines in fields as
diverse as physics, chemistry, astronomy, geoscience, biology, psychology,
material science, engineering, finance, and economics.
For example, in astronomy, NumPy was an important part of the software stack used
in the discovery of gravitational waves \cite{abbott2016observation}
and the first imaging of a black hole \cite{eht-imaging}.
% - a statement of the main conclusions (introduced by the phrase 'Here we show' or its equivalent);
Here we show how a few fundamental array concepts lead to a simple and
powerful programming paradigm for organizing, exploring, and analyzing
scientific data.
% - 2-3 sentences putting the main findings into general context so it is
%   clear how the results described in the paper have moved the field forwards.
NumPy is the foundation upon which the entire scientific Python
universe is constructed. It is so pervasive that several projects,
targeting audiences with specialized needs, have developed their own
NumPy-like interfaces and array objects. Because of its central position in the
ecosystem, NumPy increasingly plays the role of an interoperability layer
between these new array computation libraries.
